---
title: "Figure 4 script"
output: html_notebook
---

#Load necessary packages
```{r}
library(svglite)
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(ggrepel)
library(msigdbr)
library(clusterProfiler)
library(org.Mm.eg.db)
library(ggVennDiagram)
library(dplyr)
library(tibble)
library(patchwork)
library(ComplexHeatmap)
library(circlize)
library(ggh4x)
library(GOSemSim)
library(enrichplot)
library(GO.db)
library(ggrepel)
```

#Plot styling
```{r}
theme_set(
  theme_classic(base_size = 11, base_family = "Helvetica") +
    theme(
      plot.title       = element_text(size = 13, face = "plain"),
      plot.subtitle    = element_text(size = 10, color = "grey40"),
      axis.title       = element_text(size = 11),
      axis.text        = element_text(size = 10),
      legend.title     = element_text(face = "bold", size = 10),
      legend.text      = element_text(size = 9),
      legend.key.size  = unit(0.55, "lines"),
      legend.spacing.y = unit(0.15, "cm"),
      legend.position  = "right"
    )
)
plot_W <- 10
plot_H <- 8
```


#Load the data
```{r}
rna <- read_tsv(
  "~/HMS Dropbox/Daniela Coronado/Daniela's documents/Written docs/Daniela's manuscript/Manuscript version 5/Daniela final copy/Data files/Figure 4/RNASeq of BO glycolipis vs rmGal9/ZT4GNN-expression-matrix.tsv",
  show_col_types = FALSE
)

rna_clean <- rna %>%
  rename_with(~ str_replace_all(.x, " ", "")) %>%
  rename_with(tolower)

rna_long <- rna_clean %>%
  pivot_longer(
    cols         = matches("^zt4gnn_\\d+_(cpm|count)$"),
    names_to     = c("sample", "measure"),
    names_pattern = "^(zt4gnn_\\d+)_(cpm|count)$",
    values_to    = "value"
  ) %>%
  group_by(gene_id, gene_name, gene_biotype, sample, measure) %>%
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = measure, values_from = value)

```

#Organize the data
```{r}
sample_metadata <- tribble(
  ~sample,      ~BOglycolipid,       ~gal9,      ~group,
  "zt4gnn_1",   "absent",   "present",  "VHC_Gal9",
  "zt4gnn_2",   "absent",   "present",  "VHC_Gal9",
  "zt4gnn_3",   "absent",   "present",  "VHC_Gal9",
  "zt4gnn_4",   "present",  "present",  "BOglycolipid_Gal9",
  "zt4gnn_5",   "present",  "present",  "BOglycolipid_Gal9",
  "zt4gnn_6",   "present",  "present",  "BOglycolipid_Gal9",
  "zt4gnn_7",   "absent",   "absent",   "VHC_noGal9",
  "zt4gnn_8",   "absent",   "absent",   "VHC_noGal9",
  "zt4gnn_9",   "absent",   "absent",   "VHC_noGal9",
  "zt4gnn_10",  "present",  "absent",   "BOglycolipid_noGal9",
  "zt4gnn_11",  "present",  "absent",   "BOglycolipid_noGal9",
  "zt4gnn_12",  "present",  "absent",   "BOglycolipid_noGal9"
)

# Remove outlier
sample_metadata_no_outlier <- sample_metadata %>%
  filter(sample != "zt4gnn_4")

rna_long_no_outlier <- rna_long %>%
  filter(sample != "zt4gnn_4") %>%
  left_join(sample_metadata_no_outlier, by = "sample") %>%
  mutate(
    BOglycolipid   = factor(BOglycolipid,   levels = c("absent", "present")),
    gal9  = factor(gal9,  levels = c("absent", "present")),
    group = factor(group, levels = c("VHC_noGal9", "BOglycolipid_noGal9", "VHC_Gal9", "BOglycolipid_Gal9"))
  )
```

#Make count matrix for DESeq
```{r}
count_mat <- rna_long_no_outlier %>%
  dplyr::select(gene_id, sample, count) %>%
  group_by(gene_id, sample) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = sample, values_from = count, values_fill = 0) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

storage.mode(count_mat) <- "integer"

col_data <- sample_metadata_no_outlier %>%
  mutate(
    BOglycolipid   = factor(BOglycolipid,   levels = c("absent", "present")),
    gal9  = factor(gal9,  levels = c("absent", "present")),
    group = factor(group, levels = c("VHC_noGal9", "BOglycolipid_noGal9", "VHC_Gal9", "BOglycolipid_Gal9"))
  ) %>%
  column_to_rownames("sample")

col_data <- col_data[colnames(count_mat), , drop = FALSE]

dds <- DESeqDataSetFromMatrix(
  countData = count_mat,
  colData   = col_data,
  design    = ~ group
)
dds <- dds[rowSums(counts(dds)) >= 10, ]
dds <- DESeq(dds)

gene_lookup <- rna_long_no_outlier %>%
  distinct(gene_id, gene_name, gene_biotype) %>%
  mutate(across(everything(), as.character))
```


#Calculate variance stabilized counts
```{r}
# VST for heat map
vsd <- vst(dds, blind = TRUE)

vsd_mat <- assay(vsd)

id2sym   <- gene_lookup %>%
  distinct(gene_id, gene_name) %>%
  filter(!is.na(gene_name), gene_name != "")

row_syms <- id2sym$gene_name[match(rownames(vsd_mat), id2sym$gene_id)]
keep     <- !is.na(row_syms)
vsd_sym  <- vsd_mat[keep, ]
rownames(vsd_sym) <- row_syms[keep]
vsd_sym  <- vsd_sym[order(rowMeans(vsd_sym), decreasing = TRUE), ]
vsd_sym  <- vsd_sym[!duplicated(rownames(vsd_sym)), ]
```


#Build PCA plot
```{r}
pca_data    <- plotPCA(vsd, intgroup = "group", returnData = TRUE)
percent_var <- round(100 * attr(pca_data, "percentVar"))

pca_data_labeled <- pca_data %>%
  left_join(
    sample_metadata_no_outlier %>%
      mutate(group_label = case_when(
        group == "VHC_noGal9" ~ "Vehicle controls",
        group == "BOglycolipid_noGal9" ~ "BO glycolipids alone",
        group == "VHC_Gal9"   ~ "Gal-9 alone",
        group == "BOglycolipid_Gal9"   ~ "BO glycolipids + Gal-9"
      )) %>%
      dplyr::select(sample, group_label),
    by = c("name" = "sample")
  )

PCA_plot <- ggplot(
  pca_data_labeled,
  aes(x = PC1, y = PC2, color = group_label)
) +
  geom_point(size = 8, alpha = 0.9) +
  labs(
    title    = "PCA of RNA-seq samples",
    subtitle = "Variance-stabilized counts; n = 2-3",
    x        = paste0("PC1: ", percent_var[1], "% variance"),
    y        = paste0("PC2: ", percent_var[2], "% variance"),
    color    = "Group"
  )

ggsave(
  filename = "RNAseq_PCA_no_sample4.svg",
  plot     = PCA_plot,
  width    = plot_W,
  height   = plot_H,
  device   = svglite::svglite
)
```
#Precompute semantic similarity structure
```{r}
go_semdata_mm_bp <- godata(OrgDb = org.Mm.eg.db, ont = "BP")
```



#Extract top 20 GO pathways from BO glycolipid vs BO glycolipid+Gal-9
```{r}
#Extract top 20 GO pathways from BO glycolipid vs BO glycolipid+Gal-9
# Map symbols to Entrez IDs (using all gene names from gene_lookup)
symbol_to_entrez_full <- bitr(
  gene_lookup$gene_name,
  fromType = "SYMBOL",
  toType   = "ENTREZID",
  OrgDb    = org.Mm.eg.db
)

# Helper: build preranked Entrez gene list from DESeq2 result
build_entrez_ranklist <- function(res_df, stat_col, sym_col = "gene_name") {
  res_df %>%
    filter(
      !is.na(.data[[stat_col]]),
      !is.na(.data[[sym_col]]),
      .data[[sym_col]] != ""
    ) %>%
    inner_join(symbol_to_entrez_full,
               by = setNames("SYMBOL", sym_col)) %>%
    group_by(ENTREZID) %>%
    slice_max(abs(.data[[stat_col]]), n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(desc(.data[[stat_col]])) %>%
    { setNames(.[[stat_col]], .$ENTREZID) }
}

# DESeq2 results: BOglycolipid+Gal-9 vs BOglycolipid_noGal9
res_BOglycolipid_gal9_vs_BOglycolipid_nogal9 <- results(
  dds,
  contrast = c("group", "BOglycolipid_Gal9", "BOglycolipid_noGal9")
) %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  left_join(gene_lookup, by = "gene_id") %>%
  dplyr::rename(stat_BOglycolipidGal9_vs_BOglycolipid = stat)

# Preranked list by Wald statistic
gene_list_BOglycolipidGal9_vs_BOglycolipid_entrez <- build_entrez_ranklist(
  res_BOglycolipid_gal9_vs_BOglycolipid_nogal9,
  "stat_BOglycolipidGal9_vs_BOglycolipid"
)

set.seed(42)
gsea_BOglycolipidGal9_vs_BOglycolipid_go <- gseGO(
  geneList      = gene_list_BOglycolipidGal9_vs_BOglycolipid_entrez,
  OrgDb         = org.Mm.eg.db,
  ont           = "BP",
  keyType       = "ENTREZID",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  minGSSize     = 10,
  maxGSSize     = 500,
  eps           = 0,
  verbose       = FALSE
)

# Reduce redundancy
gsea_BOglycolipidGal9_vs_BOglycolipid_simplified <- simplify(
  gsea_BOglycolipidGal9_vs_BOglycolipid_go,
  cutoff     = 0.7,
  by         = "p.adjust",
  select_fun = min,
  semData    = go_semdata_mm_bp
)

# Plain data frame for downstream use
res_tbl_BOglycolipidGal9_vs_BOglycolipid <- as.data.frame(gsea_BOglycolipidGal9_vs_BOglycolipid_simplified@result)

# write full table
pathway_BOglycolipid_vs_BOglycolipidGal9 <- res_tbl_BOglycolipidGal9_vs_BOglycolipid %>%
  dplyr::arrange(p.adjust)
write.csv(
  pathway_BOglycolipid_vs_BOglycolipidGal9,
  file      = "Pathway_GSEA_BOglycolipidGal9_vs_BOglycolipidnoGal9.csv",
  row.names = FALSE
)
```

##Dot plot of top 10 positively and negatively GO pathways in BO glycolipids +/- Gal-9
```{r}

res_tbl_BOglycolipidGal9_vs_BOglycolipid <- as.data.frame(gsea_BOglycolipidGal9_vs_BOglycolipid_simplified@result)

gsea_df <- res_tbl_BOglycolipidGal9_vs_BOglycolipid %>%
  dplyr::select(dplyr::any_of(
    c("ID", "Description", "NES", "pvalue", "p.adjust", "setSize", "core_enrichment")
  )) %>%
  {
    if (!"setSize" %in% colnames(.)) {
      if ("core_enrichment" %in% colnames(.)) {
        . <- mutate(.,
          setSize = stringr::str_count(core_enrichment, "/") + 1
        )
      } else {
        . <- mutate(., setSize = NA_real_)
      }
    }
    .
  } %>%
  mutate(
    padj_plot = dplyr::case_when(
      "p.adjust" %in% colnames(.) ~ p.adjust,
      TRUE                        ~ pvalue
    )
  ) %>%
  filter(!is.na(NES), !is.na(padj_plot))

top_up <- gsea_df %>%
  filter(NES > 0) %>%
  arrange(padj_plot) %>%
  slice_head(n = 10)

top_down <- gsea_df %>%
  filter(NES < 0) %>%
  arrange(padj_plot) %>%
  slice_head(n = 10)

dot_df <- bind_rows(top_up, top_down) %>%
  mutate(neg_log10_padj = -log10(padj_plot)) %>%
  arrange(desc(NES)) %>%
  mutate(Description = factor(Description, levels = unique(Description)))

dot_plot_bidir <- ggplot(
  dot_df,
  aes(x = NES, y = Description, size = setSize, color = neg_log10_padj)
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey80") +
  geom_point(alpha = 0.9) +
  scale_color_gradient(low = "#BDD7E7", high = "#08519C",
                        name = expression(-log[10]("adjusted p-value"))) +
  scale_size_continuous(name = "Gene set size", range = c(2, 8)) +
  labs(
    x        = "Normalized Enrichment Score (NES)",
    y        = NULL,
    title    = "Top 10 positively and negatively enriched GO:BP pathways",
    subtitle = "BOglycolipid+Gal-9 vs BOglycolipid (GSEA, DESeq2 Wald statistic)\nNegative NES pathways appear at top"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.y  = element_text(size = 9),
    legend.title = element_text(size = 9),
    legend.text  = element_text(size = 8)
  )
dot_plot_bidir <- dot_plot_bidir +
  ggh4x::force_panelsizes(cols = unit(2, "in"))
dot_plot_bidir

ggsave(
  filename = "Dotplot_top10_up_down_GO_BOglycolipidGal9_vs_BOglycolipid_NESx_pvalColor_negOnTop.svg",
  plot     = dot_plot_bidir,
  width    = 8,
  height   = 6,
  device   = svglite::svglite
)
```

#Compare VHC with BO glycolipid treatment
```{r}
#Compare VHC with BO glycolipid treatment
res_BOglycolipid_vs_vhc_nogal9 <- results(
  dds,
  contrast = c("group", "BOglycolipid_noGal9", "VHC_noGal9")
) %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  left_join(gene_lookup, by = "gene_id") %>%
  dplyr::rename(stat_BOglycolipid_vs_VHC_noGal9 = stat)

gene_list_BOglycolipid_vs_VHC_noGal9_entrez <- build_entrez_ranklist(
  res_BOglycolipid_vs_vhc_nogal9,
  "stat_BOglycolipid_vs_VHC_noGal9"
)

set.seed(42)
gsea_BOglycolipid_vs_VHC_noGal9_go <- gseGO(
  geneList      = gene_list_BOglycolipid_vs_VHC_noGal9_entrez,
  OrgDb         = org.Mm.eg.db,
  ont           = "BP",
  keyType       = "ENTREZID",
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  minGSSize     = 10,
  maxGSSize     = 500,
  eps           = 0,
  verbose       = FALSE
)

# Reduce redundancy
gsea_BOglycolipid_vs_VHC_noGal9_simplified <- simplify(
  gsea_BOglycolipid_vs_VHC_noGal9_go,
  cutoff     = 0.7,
  by         = "p.adjust",
  select_fun = min,
  semData    = go_semdata_mm_bp
)

res_tbl_BOglycolipid_vs_VHC_noGal9 <- as.data.frame(gsea_BOglycolipid_vs_VHC_noGal9_simplified@result)

pathway_VHC_vs_BOglycolipid_noGal9 <- res_tbl_BOglycolipid_vs_VHC_noGal9 %>%
  dplyr::arrange(p.adjust)
write.csv(
  pathway_VHC_vs_BOglycolipid_noGal9,
  file      = "Pathway_GSEA_BOglycolipidnoGal9_vs_VHCnoGal9.csv",
  row.names = FALSE
)
```

#Dot plot for VHC vs BO glycolipid (no gal-9)
```{r}

res_tbl_BOglycolipid_vs_VHC_noGal9 <- as.data.frame(gsea_BOglycolipid_vs_VHC_noGal9_simplified@result)

gsea_df_noGal9 <- res_tbl_BOglycolipid_vs_VHC_noGal9 %>%
  dplyr::select(dplyr::any_of(
    c("ID", "Description", "NES", "pvalue", "p.adjust", "setSize", "core_enrichment")
  )) %>%
  {
    if (!"setSize" %in% colnames(.)) {
      if ("core_enrichment" %in% colnames(.)) {
        . <- mutate(.,
          setSize = stringr::str_count(core_enrichment, "/") + 1
        )
      } else {
        . <- mutate(., setSize = NA_real_)
      }
    }
    .
  } %>%
  mutate(
    padj_plot = dplyr::case_when(
      "p.adjust" %in% colnames(.) ~ p.adjust,
      TRUE                        ~ pvalue
    )
  ) %>%
  filter(!is.na(NES), !is.na(padj_plot))

top_up_noGal9 <- gsea_df_noGal9 %>%
  filter(NES > 0) %>%
  arrange(padj_plot) %>%
  slice_head(n = 10)

top_down_noGal9 <- gsea_df_noGal9 %>%
  filter(NES < 0) %>%
  arrange(padj_plot) %>%
  slice_head(n = 10)

# arrange(NES) ascending -> most positive NES becomes the LAST factor level ->
# last level plots at the TOP of a discrete y-axis in ggplot.
dot_df_noGal9 <- bind_rows(top_up_noGal9, top_down_noGal9) %>%
  mutate(neg_log10_padj = -log10(padj_plot)) %>%
  arrange(NES) %>%
  mutate(
    Description = stringr::str_wrap(Description, width = 45),
    Description = factor(Description, levels = unique(Description))
  )

dot_plot_bidir_noGal9 <- ggplot(
  dot_df_noGal9,
  aes(x = NES, y = Description, size = setSize, color = neg_log10_padj)
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey80") +
  geom_point(alpha = 0.9) +
  scale_color_gradient(low = "#BDD7E7", high = "#08519C",
                        name = expression(-log[10]("adjusted p-value"))) +
  scale_size_continuous(name = "Gene set size", range = c(2, 8)) +
  labs(
    x        = "Normalized Enrichment Score (NES)",
    y        = NULL,
    title    = "Top 10 positively and negatively enriched GO:BP pathways",
    subtitle = "BOglycolipid (no Gal-9) vs VHC (no Gal-9) (GSEA, DESeq2 Wald statistic)\nPositive NES pathways appear at top"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.y  = element_text(size = 9),
    legend.title = element_text(size = 9),
    legend.text  = element_text(size = 8),
    plot.margin  = margin(t = 10, r = 10, b = 10, l = 10),
    plot.title.position = "plot"
  ) +
  coord_cartesian(clip = "off")

dot_plot_bidir_noGal9 <- dot_plot_bidir_noGal9 +
  ggh4x::force_panelsizes(cols = unit(2, "in"))
dot_plot_bidir_noGal9

ggsave(
  filename = "Dotplot_top10_up_down_GO_BOglycolipidnoGal9_vs_VHCnoGal9_NESx_pvalColor_posOnTop.svg",
  plot     = dot_plot_bidir_noGal9,
  width    = 8,
  height   = 6,   
  device   = svglite::svglite
)
```


#Build heat map
```{r}
# Pull GO:0019882 (Antigen processing and presentation) leading edge genes from the BOglycolipid+Gal-9 vs BOglycolipid (no Gal-9) GSEA 
#Extract leading edge genes from GO:0019882

ap_go_entrez_h <- as.data.frame(gsea_BOglycolipidGal9_vs_BOglycolipid_simplified@result) %>%
  filter(ID == "GO:0019882") %>%
  pull(core_enrichment) %>%
  str_split("/") %>%
  unlist()

ap_go_symbols_h <- bitr(
  ap_go_entrez_h,
  fromType = "ENTREZID",
  toType   = "SYMBOL",
  OrgDb    = org.Mm.eg.db
)

#Filter for genes significant in the BOglycolipid+Gal-9 vs BOglycolipid contrast
sig_BOglycolipidgal9_vs_BOglycolipid <- res_BOglycolipid_gal9_vs_BOglycolipid_nogal9 %>%
  filter(!is.na(padj), padj < 0.05)

ap_go_sig_h <- ap_go_symbols_h %>%
  filter(SYMBOL %in% sig_BOglycolipidgal9_vs_BOglycolipid$gene_name) %>%
  left_join(
    sig_BOglycolipidgal9_vs_BOglycolipid %>%
      dplyr::select(gene_name, log2FoldChange, padj),
    by = c("SYMBOL" = "gene_name")
  ) %>%
  dplyr::rename(log2FC_group = log2FoldChange)

cat("Leading-edge genes passing padj < 0.05 in BOglycolipid+Gal-9 vs BOglycolipid:", nrow(ap_go_sig_h), "\n")

#Restrict to the top 20 leading-edge genes by |log2FC|
ap_go_sig_h_top20 <- ap_go_sig_h %>%
  filter(!is.na(log2FC_group)) %>%
  slice_max(order_by = abs(log2FC_group), n = 20, with_ties = FALSE)

cat("Top leading-edge genes retained for heatmap:", nrow(ap_go_sig_h_top20), "\n")

#Build z-score matrix
genes_present_go_h <- intersect(ap_go_sig_h_top20$SYMBOL, rownames(vsd_sym))

vsd_subset_go_h <- vsd_sym[genes_present_go_h, ]

zscore_mat_go_h <- t(scale(t(vsd_subset_go_h)))

# Group-averaged z-scores
group_order_go_h <- c("VHC_noGal9", "VHC_Gal9", "BOglycolipid_noGal9", "BOglycolipid_Gal9")
 
group_labels_go_h <- c(
  "VHC_noGal9" = "VHC",
  "VHC_Gal9"   = "VHC + Gal-9",
  "BOglycolipid_noGal9" = "BO glycolipid",
  "BOglycolipid_Gal9"   = "BO glycolipid + Gal-9"
)
 
zscore_group_go_h <- sapply(group_order_go_h, function(grp) {
  cols <- rownames(col_data)[col_data$group == grp]
  cols <- intersect(cols, colnames(zscore_mat_go_h))
  if (length(cols) == 1) return(zscore_mat_go_h[, cols, drop = TRUE])
  rowMeans(zscore_mat_go_h[, cols, drop = FALSE])
})
 
colnames(zscore_group_go_h) <- group_labels_go_h[group_order_go_h]
 
# Gene ordering by log2FC 
gene_order_df_go_h <- ap_go_sig_h_top20 %>%
  filter(SYMBOL %in% genes_present_go_h) %>%
  arrange(desc(log2FC_group))

# Reorder rows (genes) to match gene_order_df_go_h 
zscore_group_go_h_v <- zscore_group_go_h[gene_order_df_go_h$SYMBOL, ]
 
# olor scale 
score_lim_go_h <- quantile(abs(zscore_group_go_h_v), 0.95, na.rm = TRUE)
 
col_fun_go_h <- colorRamp2(
  c(-score_lim_go_h, 0, score_lim_go_h),
  c("#2166AC", "white", "#C0392B")
)
 
# Vertical heatmap
n_genes_h  <- nrow(zscore_group_go_h_v)
n_groups_h <- ncol(zscore_group_go_h_v)

ht_go_h <- Heatmap(
  zscore_group_go_h_v,
  name             = "Z-score",
  col              = col_fun_go_h,
  
  # Rows (genes)
  cluster_rows     = FALSE,
  row_order        = gene_order_df_go_h$SYMBOL,
  row_names_gp     = gpar(fontsize = 12),
  row_names_side   = "left",
  
  # Columns (treatment groups)
  cluster_columns  = FALSE,
  column_order     = c(
    "VHC",
    "VHC + Gal-9",
    "BO glycolipid",
    "BO glycolipid + Gal-9"
  ),
  column_names_gp  = gpar(fontsize = 12),
  column_names_rot = 45,
  column_names_max_height = unit(4, "cm"),
  
  # Borders and sizing -- equal cm per row and per column 
  rect_gp          = gpar(col = "white", lwd = 0.5),
  height           = unit(n_genes_h, "cm"),
  width            = unit(n_groups_h, "cm"),
  
  row_title        = "GO:0019882 Antigen processing and presentation\n(top 20 leading-edge genes by |log2FC|, padj < 0.05, BOglycolipid+Gal-9 vs BOglycolipid)",
  row_title_gp     = gpar(fontsize = 12)
)
 
# Export
svglite::svglite(
  filename = "AP_heatmap_GO0019882_BOglycolipidGal9_vs_BOglycolipid_top20_vertical.svg",
  width    = 12,                 # cm -- columns + row labels + legend
  height   = n_genes_h + 6        # cm -- rows + row title + rotated column labels
)
draw(
  ht_go_h,
  heatmap_legend_side = "right",
  padding = unit(c(5, 20, 5, 5), "mm")
)
dev.off()
```


#Enrichment plot
```{r}

# Confirm the term is present in the GSEA result
ap_row_check <- as.data.frame(gsea_BOglycolipidGal9_vs_BOglycolipid_simplified@result) %>%
  filter(ID == "GO:0019882")

if (nrow(ap_row_check) == 0) {
  stop(
    "GO:0019882 is not present in gsea_BOglycolipidGal9_vs_BOglycolipid_simplified -- it may ",
    "have been collapsed by simplify(). Re-run gseaplot2() on the ",
    "non-simplified object (gsea_BOglycolipidGal9_vs_BOglycolipid_go) instead."
  )
}

# Running enrichment score plot
ap_gsea_plot <- gseaplot2(
  gsea_BOglycolipidGal9_vs_BOglycolipid_simplified,
  geneSetID   = "GO:0019882",
  title       = paste0(
    "GO:0019882 Antigen processing and presentation\n",
    "BOglycolipid+Gal-9 vs BOglycolipid (NES = ", round(ap_row_check$NES, 2),
    ", padj = ", formatC(ap_row_check$p.adjust, format = "e", digits = 2), ")"
  ),
  color       = "#08519C",
  base_size   = 11,
  rel_heights = c(1.5, 0.5, 1),
  subplots    = 1:3,
  pvalue_table = FALSE
)

ap_gsea_plot

#Export
ggsave(
  filename = "GSEA_runningscore_GO0019882_BOglycolipidGal9_vs_BOglycolipid.svg",
  plot     = ap_gsea_plot,
  width    = 8,
  height   = 6,
  device   = svglite::svglite
)
```
#Scatter plot
```{r}

res_BOglycolipid_noGal9 <- res_BOglycolipid_vs_vhc_nogal9 %>%
  dplyr::rename(
    log2FC_BOglycolipid_noGal9 = log2FoldChange,
    pvalue_BOglycolipid_noGal9 = pvalue,
    padj_BOglycolipid_noGal9   = padj
  ) %>%
  dplyr::select(
    gene_id, gene_name,
    log2FC_BOglycolipid_noGal9, pvalue_BOglycolipid_noGal9, padj_BOglycolipid_noGal9
  )
 
# BOglycolipid_Gal9 vs VHC_Gal9
res_BOglycolipid_Gal9 <- results(
  dds,
  contrast = c("group", "BOglycolipid_Gal9", "VHC_Gal9")
) %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  left_join(gene_lookup, by = "gene_id") %>%
  dplyr::rename(
    log2FC_BOglycolipid_Gal9 = log2FoldChange,
    pvalue_BOglycolipid_Gal9 = pvalue,
    padj_BOglycolipid_Gal9   = padj
  ) %>%
  dplyr::select(
    gene_id,
    log2FC_BOglycolipid_Gal9, pvalue_BOglycolipid_Gal9, padj_BOglycolipid_Gal9
  )
 
#Merge contrasts into one scatter-ready table
de_scatter <- res_BOglycolipid_noGal9 %>%
  left_join(res_BOglycolipid_Gal9, by = "gene_id")
 
# Filter to genes significant (padj < 0.05, |log2FC| > 1.5) in at least one of the two contrasts 
de_scatter_sig <- de_scatter %>%
  filter(
    (padj_BOglycolipid_noGal9 < 0.05 & abs(log2FC_BOglycolipid_noGal9) > 1.5) |
      (padj_BOglycolipid_Gal9 < 0.05 & abs(log2FC_BOglycolipid_Gal9) > 1.5)
  )
 
# Classify directionality of the BOglycolipid effect with vs without Gal-9 
de_scatter_sig <- de_scatter_sig %>%
  mutate(
    plot_color = case_when(
 
      # ── Top-right: both > 1.5, split by diagonal ──────────────────────────
      log2FC_BOglycolipid_noGal9 > 1.5 & log2FC_BOglycolipid_Gal9 > 1.5 & log2FC_BOglycolipid_Gal9 > log2FC_BOglycolipid_noGal9 ~
        "Stronger with Gal-9",
      log2FC_BOglycolipid_noGal9 > 1.5 & log2FC_BOglycolipid_Gal9 > 1.5 & log2FC_BOglycolipid_Gal9 < log2FC_BOglycolipid_noGal9 ~
        "Stronger without Gal-9",
      log2FC_BOglycolipid_noGal9 > 1.5 & log2FC_BOglycolipid_Gal9 > 1.5 ~
        "Equal (up)",
 
      # ── Top-right extensions ───────────────────────────────────────────────
      log2FC_BOglycolipid_noGal9 > 0 & log2FC_BOglycolipid_noGal9 < 1.5 & log2FC_BOglycolipid_Gal9 > 1.5 ~
        "Stronger with Gal-9",
      log2FC_BOglycolipid_noGal9 > 1.5 & log2FC_BOglycolipid_Gal9 > 0 & log2FC_BOglycolipid_Gal9 < 1.5 ~
        "Stronger without Gal-9",
 
      # ── Bottom-left: both < -1.5, split by diagonal ───────────────────────
      log2FC_BOglycolipid_noGal9 < -1.5 & log2FC_BOglycolipid_Gal9 < -1.5 & log2FC_BOglycolipid_Gal9 < log2FC_BOglycolipid_noGal9 ~
        "Suppression stronger with Gal-9",
      log2FC_BOglycolipid_noGal9 < -1.5 & log2FC_BOglycolipid_Gal9 < -1.5 & log2FC_BOglycolipid_Gal9 > log2FC_BOglycolipid_noGal9 ~
        "Suppression stronger without Gal-9",
      log2FC_BOglycolipid_noGal9 < -1.5 & log2FC_BOglycolipid_Gal9 < -1.5 ~
        "Equal (down)",
 
      # ── Bottom-left extensions ─────────────────────────────────────────────
      log2FC_BOglycolipid_noGal9 < 0 & log2FC_BOglycolipid_noGal9 > -1.5 & log2FC_BOglycolipid_Gal9 < -1.5 ~
        "Suppression stronger with Gal-9",
      log2FC_BOglycolipid_noGal9 < -1.5 & log2FC_BOglycolipid_Gal9 < 0 & log2FC_BOglycolipid_Gal9 > -1.5 ~
        "Suppression stronger without Gal-9",
 
      TRUE ~ "Other"
    )
  )
 
# Plot styling inputs
plot_colors <- c(
  "Stronger with Gal-9"                = "#2166AC",
  "Stronger without Gal-9"             = "#C0392B",
  "Equal (up)"                         = "grey60",
  "Suppression stronger with Gal-9"    = "#6A0DAD",
  "Suppression stronger without Gal-9" = "#1B7A4A",
  "Equal (down)"                       = "grey60",
  "Other"                              = "grey85"
)
 
# Panel size for force_panelsizes() 
panel_size <- unit(4, "in")
 
# Shading polygons 
ax_max <- ceiling(max(
  abs(de_scatter_sig$log2FC_BOglycolipid_noGal9),
  abs(de_scatter_sig$log2FC_BOglycolipid_Gal9),
  na.rm = TRUE
))
 
# Blue: rectangle (0 < x < 1.5, y > 1.5) + triangle (x > 1.5, y > x)
tr_blue <- data.frame(
  x = c(0,      1.5,    ax_max, 0     ),
  y = c(1.5,    1.5,    ax_max, ax_max)
)
 
# Red: triangle (y > 1.5, x > y) + rectangle (x > 1.5, 0 < y < 1.5)
tr_red <- data.frame(
  x = c(1.5,    ax_max, ax_max, 1.5),
  y = c(0,      0,      ax_max, 1.5)
)
 
# Purple: rectangle (-1.5 < x < 0, y < -1.5) + triangle (x < -1.5, y < x)
bl_purple <- data.frame(
  x = c(0,      -1.5,    -ax_max, 0      ),
  y = c(-1.5,   -1.5,    -ax_max, -ax_max)
)
 
# Green: triangle (y < -1.5, x < y) + rectangle (x < -1.5, -1.5 < y < 0)
bl_green <- data.frame(
  x = c(-1.5,   -ax_max, -ax_max, -1.5),
  y = c(0,       0,      -ax_max, -1.5)
)
 
# Plot 
scatter_plot_1 <- ggplot(de_scatter_sig, aes(x = log2FC_BOglycolipid_noGal9, y = log2FC_BOglycolipid_Gal9)) +
 
  # Shaded regions
  geom_polygon(data = tr_blue,   aes(x = x, y = y),
               fill = "#2166AC", alpha = 0.06, inherit.aes = FALSE) +
  geom_polygon(data = tr_red,    aes(x = x, y = y),
               fill = "#C0392B", alpha = 0.06, inherit.aes = FALSE) +
  geom_polygon(data = bl_purple, aes(x = x, y = y),
               fill = "#6A0DAD", alpha = 0.06, inherit.aes = FALSE) +
  geom_polygon(data = bl_green,  aes(x = x, y = y),
               fill = "#1B7A4A", alpha = 0.06, inherit.aes = FALSE) +
 
  # Gray points first (bottom layer)
  geom_point(
    data  = ~filter(.x, plot_color == "Other"),
    color = "grey85",
    size  = 1.8, alpha = 0.6
  ) +
  # Colored points on top
  geom_point(
    data = ~filter(.x, plot_color != "Other"),
    aes(color = plot_color),
    size = 1.8, alpha = 0.75
  ) +
 
  # Reference lines
  geom_hline(yintercept = 0,    linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_vline(xintercept = 0,    linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_hline(yintercept =  1.5, linetype = "dashed", linewidth = 0.3, color = "grey70") +
  geom_hline(yintercept = -1.5, linetype = "dashed", linewidth = 0.3, color = "grey70") +
  geom_vline(xintercept =  1.5, linetype = "dashed", linewidth = 0.3, color = "grey70") +
  geom_vline(xintercept = -1.5, linetype = "dashed", linewidth = 0.3, color = "grey70") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", linewidth = 0.7, color = "grey30") +
 
  scale_color_manual(
    values = plot_colors,
    labels = c(
      "Stronger with Gal-9"                = "Upregulation stronger with Gal-9",
      "Stronger without Gal-9"             = "Upregulation stronger without Gal-9",
      "Equal (up)"                         = "Upregulation equal with/without Gal-9",
      "Suppression stronger with Gal-9"    = "Suppression stronger with Gal-9",
      "Suppression stronger without Gal-9" = "Suppression stronger without Gal-9",
      "Equal (down)"                       = "Suppression equal with/without Gal-9"
    ),
    na.translate = FALSE
  ) +
  coord_fixed(xlim = c(-ax_max, ax_max), ylim = c(-ax_max, ax_max)) +
  guides(
    color = guide_legend(override.aes = list(size = 4, alpha = 1), ncol = 1)
  ) +
  labs(
    title    = "DC response to BO glycolipid with and without exogenous Gal-9",
    subtitle = "Genes shown: adjusted p-value < 0.05 and absolute log2 fold-change (FC) > 1.5 in at least one contrast",
    x        = "BO glycolipid effect without Gal-9 (log2FC)",
    y        = "BO glycolipid effect with Gal-9 (log2FC)",
    color    = "Response class"
  ) +
  theme(
    legend.box.margin = margin(0, 0, 0, 6)
  ) +
  ggh4x::force_panelsizes(rows = panel_size, cols = panel_size)
 
scatter_plot_1
 
ggsave(
  filename = "BOglycolipid_response_scatter_diagonal_shading_no_sample4.svg",
  plot     = scatter_plot_1,
  width    = plot_W,
  height   = plot_H,
  device   = svglite::svglite
)
```

#Overlay genes into scatterplot

```{r}

#Pull every descendant GO:BP term of GO:0019882 
ap_offspring <- as.list(GOBPOFFSPRING["GO:0019882"])[["GO:0019882"]]
ap_all_ids   <- c(ap_offspring)

ap_terms <- AnnotationDbi::select(
  GO.db,
  keys    = ap_all_ids,
  keytype = "GOID",
  columns = "TERM"
)

# Test child terms directly using the NON-simplified GSEA object. gsea_BOglycolipidGal9_vs_BOglycolipid_go (pre-simplify()), not gsea_BOglycolipidGal9_vs_BOglycolipid_simplified. 
ap_child_gsea <- as.data.frame(gsea_BOglycolipidGal9_vs_BOglycolipid_go@result) %>%
  filter(ID %in% ap_all_ids) %>%
  left_join(ap_terms, by = c("ID" = "GOID"))

# Recompute FDR WITHIN just this subtree, not genome-wide. 
ap_child_gsea <- ap_child_gsea %>%
  mutate(p.adjust_subtree = p.adjust(pvalue, method = "BH")) %>%
  arrange(p.adjust_subtree)

# Report 
cat("Child terms of GO:0019882 actually tested by gseGO():", nrow(ap_child_gsea), "\n")
cat("Child terms significant WITHIN the subtree (p.adjust_subtree < 0.05):",
    sum(ap_child_gsea$p.adjust_subtree < 0.05), "\n\n")

ap_child_gsea_report <- ap_child_gsea %>%
  dplyr::select(ID, TERM, NES, setSize, pvalue, p.adjust_subtree)

print(ap_child_gsea_report)
```


#MHCII and MHCI pathways enriched within the parent pathway 
#Highlight top 5 leading edge genes for each
```{r}
# Define the GO IDs backing each category, based on the subtree. GSEA results table (child terms tested directly, non-simplified object, subtree FDR-corrected) 
mhc2_ids <- c(
  "GO:0002495",  # antigen processing and presentation of peptide antigen via MHC class II
  "GO:0019886",  # ...of exogenous peptide antigen via MHC class II
  "GO:0002504",  # ...of peptide or polysaccharide antigen via MHC class II
  "GO:0002503"   # peptide antigen assembly with MHC class II protein complex
)

mhc1_ids <- c(
  "GO:0042590",  # antigen processing and presentation of exogenous peptide antigen via MHC class I (cross-presentation)
  "GO:0002474"   # antigen processing and presentation of peptide antigen via MHC class I
)

# Pull leading-edge genes for each 
extract_leading_edge <- function(gsea_obj, go_ids) {
  as.data.frame(gsea_obj@result) %>%
    filter(ID %in% go_ids) %>%
    pull(core_enrichment) %>%
    str_split("/") %>%
    unlist() %>%
    unique()
}

mhc2_entrez <- extract_leading_edge(gsea_BOglycolipidGal9_vs_BOglycolipid_go, mhc2_ids)
mhc1_entrez <- extract_leading_edge(gsea_BOglycolipidGal9_vs_BOglycolipid_go, mhc1_ids)

cat("Unique leading-edge genes -- MHC class II terms:", length(mhc2_entrez), "\n")
cat("Unique leading-edge genes -- MHC class I terms:", length(mhc1_entrez), "\n")

# Convert to gene symbols
mhc2_symbols <- bitr(mhc2_entrez, fromType = "ENTREZID", toType = "SYMBOL", OrgDb = org.Mm.eg.db)
mhc1_symbols <- bitr(mhc1_entrez, fromType = "ENTREZID", toType = "SYMBOL", OrgDb = org.Mm.eg.db)

# Rank by EFFECT SIZE, restricted to genes that are significant. First filter to padj < 0.05 (must be significant), THEN rank by |log2FoldChange| 
mhc2_top5 <- res_BOglycolipid_gal9_vs_BOglycolipid_nogal9 %>%
  filter(gene_name %in% mhc2_symbols$SYMBOL, !is.na(padj), padj < 0.05) %>%
  slice_max(order_by = abs(log2FoldChange), n = 5, with_ties = FALSE) %>%
  mutate(mhc_category = "MHC class II")

mhc1_top5 <- res_BOglycolipid_gal9_vs_BOglycolipid_nogal9 %>%
  filter(gene_name %in% mhc1_symbols$SYMBOL, !is.na(padj), padj < 0.05) %>%
  slice_max(order_by = abs(log2FoldChange), n = 5, with_ties = FALSE) %>%
  mutate(mhc_category = "MHC class I")

# Combined highlight table 
mhc_top5_highlight <- bind_rows(mhc1_top5, mhc2_top5) %>%
  dplyr::select(gene_name, mhc_category, log2FoldChange, padj, stat_BOglycolipidGal9_vs_BOglycolipid)

cat("\nTop 5 MHC class I genes (BOglycolipid+Gal-9 vs BOglycolipid):\n")
print(mhc1_top5 %>% dplyr::select(gene_name, log2FoldChange, padj))

cat("\nTop 5 MHC class II genes (BOglycolipid+Gal-9 vs BOglycolipid):\n")
print(mhc2_top5 %>% dplyr::select(gene_name, log2FoldChange, padj))

```

#Build scatter plot highlighting what's above
```{r}

# Join the highlight genes onto the scatter coordinates 
mhc_highlight_coords <- mhc_top5_highlight %>%
  dplyr::select(gene_name, mhc_category) %>%
  inner_join(
    de_scatter_sig %>%
      dplyr::select(gene_name, log2FC_BOglycolipid_noGal9, log2FC_BOglycolipid_Gal9),
    by = "gene_name"
  )

cat("MHC highlight genes with plottable coordinates:", nrow(mhc_highlight_coords), "\n")
print(mhc_highlight_coords)

n_expected <- nrow(mhc_top5_highlight)
n_matched  <- nrow(mhc_highlight_coords)
if (n_matched < n_expected) {
  cat(
    "\nNOTE:", n_expected - n_matched,
    "highlight gene(s) were not found in de_scatter_sig and will not appear",
    "on the plot -- check mhc_top5_highlight vs de_scatter_sig$gene_name.\n"
  )
}

# Colors for the two highlighted categories
mhc_highlight_colors <- c(
  "MHC class I"  = "#1F78B4",
  "MHC class II" = "#33A02C"
)

# Scatter plot 
scatter_plot_3 <- ggplot(de_scatter_sig, aes(x = log2FC_BOglycolipid_noGal9, y = log2FC_BOglycolipid_Gal9)) +

# Quadrant shading (reused from scatter_plot_1)
  geom_polygon(data = tr_blue,   aes(x = x, y = y),
               fill = "#2166AC", alpha = 0.06, inherit.aes = FALSE) +
  geom_polygon(data = tr_red,    aes(x = x, y = y),
               fill = "#C0392B", alpha = 0.06, inherit.aes = FALSE) +
  geom_polygon(data = bl_purple, aes(x = x, y = y),
               fill = "#6A0DAD", alpha = 0.06, inherit.aes = FALSE) +
  geom_polygon(data = bl_green,  aes(x = x, y = y),
               fill = "#1B7A4A", alpha = 0.06, inherit.aes = FALSE) +
 
  # Reference lines
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_hline(yintercept =  1.5, linetype = "dashed", linewidth = 0.3, color = "grey70") +
  geom_hline(yintercept = -1.5, linetype = "dashed", linewidth = 0.3, color = "grey70") +
  geom_vline(xintercept =  1.5, linetype = "dashed", linewidth = 0.3, color = "grey70") +
  geom_vline(xintercept = -1.5, linetype = "dashed", linewidth = 0.3, color = "grey70") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", linewidth = 0.7, color = "grey30") +
 
  # All significant genes as light grey background
  geom_point(color = "grey85", size = 1.6, alpha = 0.5) +
 
  # Highlighted MHC-I / MHC-II genes on top
   geom_point(
    data = mhc_highlight_coords,
    aes(x = log2FC_BOglycolipid_noGal9, y = log2FC_BOglycolipid_Gal9, fill = mhc_category),
    shape  = 21,          # filled circle with a separate border color
    color  = "black",     # border color
    stroke = 0.4,          # border thickness
    size   = 2, alpha = 0.95
  ) +
 
  # Gene labels for the highlighted set only
  ggrepel::geom_text_repel(
    data = mhc_highlight_coords,
    aes(x = log2FC_BOglycolipid_noGal9, y = log2FC_BOglycolipid_Gal9, label = gene_name),
    size                = 3,
    fontface            = "italic",
    max.overlaps        = 20,
    segment.size        = 0.3,
    segment.color       = "grey40",
    min.segment.length  = 0,
    box.padding         = 0.4
  ) +
 
  scale_color_manual(
    values = mhc_highlight_colors,
    name   = "Top 5 significant genes\nby |log2FC| (Gal-9 added)"
  ) +
  coord_fixed(xlim = c(-ax_max, ax_max), ylim = c(-ax_max, ax_max)) +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1), ncol = 1)) +
  labs(
    title    = "Top 5 MHC class I and MHC class II genes driving the Gal-9 effect",
    subtitle = "Ranked by |log2FC| among genes significant (padj < 0.05) in BOglycolipid+Gal-9 vs BOglycolipid",
    x        = "BO glycolipid effect without Gal-9 (log2FC)",
    y        = "BO glycolipid effect with Gal-9 (log2FC)"
  ) +
  theme(
    legend.box.margin = margin(0, 0, 0, 6)
  ) +
  ggh4x::force_panelsizes(rows = panel_size, cols = panel_size)
 
scatter_plot_3
 
ggsave(
  filename = "BOglycolipid_response_scatter_MHC_top5_highlight.svg",
  plot     = scatter_plot_3,
  width    = plot_W,
  height   = plot_H,
  device   = svglite::svglite
)
```


